---
output: github_document
bibliography: "inst/REFERENCES.bib"
csl: "inst/apa-6th.csl"
---

# ANOPA: Analysis of Proportions using Anscombe transform

<!-- badges: start -->
[![CRAN Status](https://www.r-pkg.org/badges/version/ANOPA)](https://cran.r-project.org/package=ANOPA)
<!-- badges: end -->

```{r, echo = FALSE, message = FALSE, results = 'hide', warning = FALSE}
cat("this will be hidden; used for general initializations.\n")
library(ANOPA)
options("ANOPA.feedback" = "none") # shut down all information
```

The library `ANOPA` provides easy-to-use tools to analyze proportions . 
With it, you can examine
if proportions are significantly different (_show an effect_). In the
case where there is more than one factor, you can also test if the
interaction(s) are significant. You 
can also test simple effects (a.k.a. _expected marginal_ analysis),
as well as post-hoc tests (using Tukey's _Honestly Significant Difference_ test HSD). 
Finally, you can assess differences based on orthogonal contrasts.
You can consult @lc23 for details.

ANOPA also comes (a) with tools to make a plot of the proportions along
with 95% confidence intervals [these intervals are adjusted for pair-
wise comparisons; @cgh21]; (b) with tools to compute statistical power given
some _a priori_ expected proportions or sample size to reach a certain
statistical power; (c) to generate random proportions if you wish to perform
Monte Carlo simulations on proportions. 
In sum, eveything you need to analyse proportions!

The main function is `anopa()` which returns an omnibus analysis of the 
proportions for the factors given. For example, if you have a data frame
`ArticleExample2` which contains a column called `s` where the 
number of successes per group are stored,
and a column called `n` where the group sizes are stored, then the following
performs an analysis of proportions as a function of the groups based on the
columns `SES` and `MofDiagnostic`:

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
w <- anopa( {s; n} ~ SES * MofDiagnostic, ArticleExample2 )
summary(w)
```

As the results suggest (consult the first three columns), there is
a main effect of the factor SES (F(2, inf) = 6.395, p = .002).
A plot of the proportions can be obtained easily with 

```{r, message=FALSE, warning=FALSE, fig.width=5.25, fig.height=3}
anopaPlot(w) 
```

or just the main effect figure with 

```{r, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
anopaPlot(w, ~ SES)
```

If the interaction had been significant, simple effects can be analyzed from the _expected marginal
frequencies_ with `e <- emProportions(w, ~ SES | MofDiagnostic )`.

Follow-up analyses include contrasts examinations with `contrastProportions()`; finally,
post-hoc pairwise comparisons can be obtained with `posthocProportions()`.


Prior to running an experiment, you might consider some statistical power planning 
on proportions using ``anopaPower2N()`` or
``anopaN2Power()`` as long as you can anticipate the expected proportions. A 
convenient effect size, the f-square and eta-square can be obtained with `anopaPropTofsq()`.

Finally, `toCompiled()`, `toLong()` and `toWide()` 
can be used to present the proportion in other formats.

# Installation

The official **CRAN** version can be installed with 

```{r, echo = TRUE, eval = FALSE}
install.packages("ANOPA")
library(ANOPA)
```

The development version `r packageVersion("ANOPA")` can be accessed through GitHub:

```{r, echo = TRUE, eval = FALSE}
devtools::install_github("dcousin3/ANOPA")
library(ANOPA)
```

Note that the package `ANOPA` is named using UPPERCASE letters whereas the main function 
`anopa()` is written using lowercase letters.

The library is loaded with 

```{r, echo = TRUE, eval = FALSE, results = FALSE}
library(ANOPA)
```

# In sum

As seen, the library `ANOPA` makes it easy to analyze proportions using the
same general vocabulary found in ANOVAs.

The complete documentation is available on this 
[site](https://dcousin3.github.io/ANOPA/).

A general introduction to the `ANOPA` framework underlying this 
library can be found at @lc23.

# References

\insertAllCited{}
