---
title: "Data formats for proportions"
bibliography: "../inst/REFERENCES.bib"
csl: "../inst/apa-6th.csl"
output: 
  rmarkdown::html_vignette
description: >
  This vignette describes the various ways that proportions can be entered in a data.frame.
vignette: >
  %\VignetteIndexEntry{Data formats for proportions}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE, results = 'hide', warning = FALSE}
cat("this is hidden; general initializations.\n")
library(ANOPA)
w1 <- anopa( {success;total} ~ Class * Difficulty, twoWayExample)
dataWide1     <- toWide(w1)
dataCompiled1 <-toCompiled(w1)
dataLong1     <- toLong(w1)
rownames(dataLong1) <- NULL

w2 <- anopa( cbind(bpre, bpost, b1week, b5week) ~ Status, minimalMxExample, WSFactors = "Moment(4)")
dataWide2     <- toWide(w2)
dataCompiled2 <-toCompiled(w2)
dataLong2     <- toLong(w2)
rownames(dataLong2) <- NULL
```



# Data formats for proportions

proportions are actually not raw data: they are the proportion of one response 
(typically called a `success`) over all the responses (the other responses being
called collectively a `failure`). As such, a proportion is a _summary statistic_, 
a bit like the mean is a summary statistic of continuous data. 

Very often, the `success` are coded using the digit `1` and the `failure`, with 
the digit `0`. When this is the case, computing the mean is actually the same
as computing the proportion of successes. However, it is a conceptual mistake to 
think of proportions as means, because they must the processed completely differently 
from averages. For example, standard error and confidence intervals for proportions
are obtained using very different procedures than standard error and confidence intervals 
for the mean.

In this vignette, we review various ways that data can be coded in a data frame. 
In a nutshell, there are three ways to represent success or failures, Wide, 
Long, and Compiled. The first two shows raw scores whereas the last shows 
a summary of the data.

Before we begin, we load the package ``ANOPA``
(if is not present on your computer, first upload it to your computer from 
CRAN or from the source repository
``devtools::install_github("dcousin3/ANOPA")``): 

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
library(ANOPA)   
```



## First format: Wide data format

In this format, there is one line per _subject_ and one column for each 
measurements. The columns contain only 1s (`success`) or 0s (`failure). 

If the particpant was measured multiple times, there is one (or some) within-subject
factor(s) resulting in multiple columns of measurements. In between-group design,
there is only a single column of scores.

As an example, consider the following data for a between-subject factor design
with two factors: Class (2 levels) and Difficulty (3 levels) for 6 groups. There
is an identical number of participants in each, 12, for a total of 72 participants.

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
dataWide1
```

When the data are in a wide format, the formula in ``anopa()`` must
provide the columns where the success/failure are stored, and the conditions
after the usual ~, as in 

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
w1 <- anopa( success ~ Class * Difficulty, dataWide1)
```

(how dataWide1 was obtained is shown below in the Section 
*Converting between formats* below.)

As another example, consider the following example obtained in a mixed, within- and between-
subject design. It has a factor `Status` with 8, 9 and 7 participants per group respectively. It 
also has four repeated measures, `bpre`, `bpost`, `b1week` and `b5week` which represent
four different Moments of measurements. The data frame is

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
dataWide2
```

The formula for analyzing these data in this format is 

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
w2 <- anopa( cbind(bpre, bpost, b1week, b5week) ~ Status, dataWide2, WSFactors = "Moment(4)" )
```

It is necessary to (a) group all the measurement columns using `cbind()`;
(b) indicate the within-subject factor(s) using the argument `WSFactors`
along with the number of levels each in a string.




## Second format: Long data format

This format may be prefered for linear modelers (but it may rapidly becomes
_very_ long!). There is always at least these columns: One Id column, one
column to indicate a within-subject level, and one column to indicate the observed 
score.  On the other hand, this format has fewer columns in repeated measure designs.

This example shows the first 6 lines of the 2-factor between design data above, stored in the long format.

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE}
head(dataLong1)
```

To analyse such data format within ``anopa()``, use

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
w1Long <- anopa( Value ~ Class * Difficulty * Variable  | Id, dataLong1 )
```

The vertical line symbol indicates that the observations are nested within
``Id`` (i.e., all the lines with the same Id are actually the same subject).


With the mixed design described above, the data begin as:

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
head(dataLong2)
```

and are analyzed with the formula:

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
w2Long <- anopa( Value ~ Status * Variable  | Id, dataLong2, WSFactors="Moment(4)" )
```


## Third format: Compiled data format

This format is compiled, in the sense that the 0s and 1s have been
replaced by a single count of success for each cell of the design. 
Hence, we no longer have access to the raw data. This format 
however has the advantage of being very compact, requiring
few lines. Here is the data for the 2 between-subject factors example

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE}
dataCompiled1
```

To use a compiled format in `anopa()`, use

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
w1Compiled <- anopa( {success; Count} ~ Class * Difficulty, dataCompiled1 )
```

where ``succes`` identifies in which column the total number of successes 
are stored. The column Count indicates the total number of observations in 
that cell. The notation {s;n} is read ``s over n`` (note the curly braces and semicolon).

For the mixed design presented earlier, the data looks like:

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE}
dataCompiled2
```

where there are columns for the number of success for each repeated measures. A new
columns appear ``uAlpha``. This column (called _unitary alpha_) is a measure of 
correlation (between -1 and +1). In this ficticious example, the correlations are
near zero (negative actually) by chance as the data were generated randomly.

It is not possible to run an ANOPA analysis at this time on compiled data
when there are repeated measures (but this may change in a future version).



## Converting between formats

Once entered in an ``anopa()`` structure, it is possible to 
convert to any format using ``toWide()``, ``toCompiled()``
and ``toLong()``. For example:

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
toCompiled(w1)
toCompiled(w2)
```

The compiled format is probably the most compact format, but the 
wide format is the most explicite format (as we see all the subjects and
their scores on a single line, one subject per line).

## Getting the example data frame

Above, we used two examples. They are available in this package under the 
names ``twoWayExample`` and ``minimalMxExample``. The first is available 
in compiled form, the second in wide form.

We converted these data set in other formats using:

```{r}
w1 <- anopa( {success;total} ~ Class * Difficulty, twoWayExample)
dataWide1     <- toWide(w1)
dataCompiled1 <-toCompiled(w1)
dataLong1     <- toLong(w1)

w2 <- anopa( cbind(bpre, bpost, b1week, b5week) ~ Status, minimalMxExample, WSFactors = "Moment(4)")
dataWide2     <- toWide(w2)
dataCompiled2 <-toCompiled(w2)
dataLong2     <- toLong(w2)
```

## Multiple repeated-measure factors

One limitation is with regards to repeated measures: It is not possible
to guess the name of the within-subject factors from the names of the columns. 
This is why, as soon as there are more than one measurement, the argument
``WSFactors`` must be added. 

Suppose a two-way within-subject design with 2 x 3 levels. The
data set ``twoWayWithinExample`` has 6 columns; the first three are
for the factor A, level 1, and the last three are for factor A, level 2.
Within each triplet of column, the factor B goes from 1 to 3.


```{r, message=TRUE, warning=FALSE, echo=TRUE, eval=TRUE}
w3 <- anopa( cbind(r11,r12,r12,r21,r22,r23) ~ . , 
             twoWayWithinExample, 
             WSFactors = c("A(3)","B(2)") 
            )
toCompiled(w3)
```

A "fyi" message is shown which lets you see how the variables are interpreted.


To know more about analyzing proportions with ANOPA, refer to @lc23 or to
[What is an ANOPA?](../articles/A-WhatIsANOPA.html).



# References





