---
title: "Testing type-I error rates"
bibliography: "../inst/REFERENCES.bib"
csl: "../inst/apa-6th.csl"
output: 
  rmarkdown::html_vignette
description: >
  This vignette describes how to test type-I error rates in the ANOPA.
vignette: >
  %\VignetteIndexEntry{Testing type-I error rates}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE, results = 'hide', warning = FALSE}
cat("this is hidden; general initializations.\n")

```


**A valid statistical test is one where the amount of type-I errors
(rejecting the null when it should not) does not exceed the test
threshold (often 5%).** We say it in bold because it seems to be commonly
forgotten. Logistic regressions applied to proportions deviates massively from 
this rule, becoming very liberal tests.

ANOPA respect very closely this rule when the corrected statistics are 
consulted.

The following code shows how to test the type-I error rate for ANOPA.It uses simulated
data with no differences. The only limitations herein is that no
correlations is added when repeated measures are used.

We suggests here to shut down error and feedback messages:

```{r, warning=FALSE, message=FALSE}
options("ANOPA.feedback" = 'none')
library(ANOPA)
library(testthat)
nsim   <- 1000     # increase for more reliable simulations.
theN   <- 20       # number of simulated participants
```

Note that the simulations are actually not run in this vignette, as 
they take times. We wished to provide code in case you wished to test
type-I error rate by yourself. The present code is also not optimized 
for speed (and in particular, is not parallelized); we wished to 
keep the code as simple as possible for readability. In all the following,
the number of simulated participants per group is small (20) but can be
varied.


# Simulations with a single factor

## Simulation with one between factor

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
frm      <- s ~ grp         # the formula
BSDesign <- list(grp = c("ctrl","plcbo")) #one factor, two groups
thePs    <- c(0.3, 0.3)     # the true proportions, equal

# test type-I error rate when no effect as is the case for factor 2
set.seed(41)
res <- c()
for (i in 1:nsim) {
    smp <- GRP( thePs, theN, BSDesign )
    w   <- anopa(frm, smp[,2:3] )
    res <- c(res, if(summarize(w)[1,4]<.05) 1 else 0)
}
typeI <- mean(res)
cat( "Design B,     testing B: ", typeI, "\n")

# tolerance is large as the number of simulations is small
expect_equal( typeI, .05, tolerance = 0.035) 
```


## Simulation with one within factor

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
frm      <- cbind(s.early, s.middle, s.late) ~ .
WSDesign <- list(moment = c("early","middle","late"))
thePs    <- c(0.3, 0.3, 0.3)

# test type-I error rate when no effect as is the case for factor 2
set.seed(42)
res <- c()
for (i in 1:nsim) {
    smp <- GRP( thePs, theN, NULL, WSDesign )
    w   <- anopa(frm, smp[,2:4] , WSFactors = "M(3)" )
    res <- c(res, if(summarize(w)[1,4]<.05) 1 else 0)
}
typeI <- mean(res)
cat( "Design W,     testing W: ", typeI, "\n")

# tolerance is large as the number of simulations is small
expect_equal( typeI, .05, tolerance = 0.035)
```


# Simulations with two factors

## Simulation with two factors, between design

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
frm      <- s ~ grp * eta
WSDesign <- list()
BSDesign <- list(eta = c("repue","ajun"), grp = c("early","middle","late"))
thePs    <- c(0.3, 0.3, 0.5, 0.5, 0.7, 0.7)

# test type-I error rate when no effect as is the case for factor 2
set.seed(41)
res <- c()
for (i in 1:nsim) {
    smp <- GRP( thePs, theN, BSDesign )
    w   <- anopa(frm, smp )
    res <- c(res, if(summarize(w)[2,4]<.05) 1 else 0)
}
typeI <- mean(res)
cat( "Design BxB,   testing B: ", typeI, "\n")

# tolerance is large as the number of simulations is small
expect_equal( typeI, .05, tolerance = 0.035)
```


## Simulation with two factors, within design

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
frm      <- cbind(s.repue.early, s.ajun.early, 
                  s.repue.middle, s.ajun.middle, 
                  s.repue.late, s.ajun.late) ~ .
BSDesign <- list()
WSDesign <- list(eta = c("repue","ajun"), moment = c("early","middle","late"))
thePs    <- c(0.3, 0.3, 0.5, 0.5, 0.7, 0.7)
# thePs    <- c(0.3, 0.7, 0.3, 0.7, 0.3, 0.7) # or no effect on factor 1

# test type-I error rate when no effect as is the case for factor 2
set.seed(41)
res <- c()
for (i in 1:nsim) {
    smp <- GRP( thePs, theN, NULL, WSDesign )
    w   <- anopa(frm, smp, WSFactors = c("e(2)", "m(3)") )
    res <- c(res, if(summarize(w)[2,4]<.05) 1 else 0)
}
typeI <- mean(res)
cat( "Design WxW,   testing W: ", typeI, "\n")

# tolerance is large as the number of simulations is small
expect_equal( typeI, .05, tolerance = 0.035)
```



## Simulation with two factors, mixed design

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
frm      <- cbind(s.early, s.middle, s.late) ~ grp
BSDesign <- list(grp = c("ctrl","plcbo"))
WSDesign <- list(moment = c("early","middle","late"))
thePs    <- c(0.3, 0.3, 0.5, 0.5, 0.7, 0.7)
# thePs    <- c(0.3, 0.7, 0.3, 0.7, 0.3, 0.7) # or no effect on factor 1

# test type-I error rate when no effect as is the case for factor 2
set.seed(41)
res <- c()
for (i in 1:nsim) {
    smp <- GRP( thePs, theN, BSDesign, WSDesign )
    w   <- anopa(frm, smp[,2:5] , WSFactors = "M(3)")
    res <- c(res, if(summarize(w)[2,4]<.05) 1 else 0)
}
typeI <- mean(res)
cat( "Design WxB,   testing B: ", typeI, "\n")

# tolerance is large as the number of simulations is small
expect_equal( typeI, .05, tolerance = 0.035)
```



# Simulations with three factors

## Simulation with three factors, all between design

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
frm      <- s ~ grp * eta * a
BSDesign <- list(eta = c("repue","ajun"), 
                 grp = c("early","middle","late"), a = c("1","2","3","4"))
thePs    <- rep(0.3, 24)

# test type-I error rate when no effect as is the case for factor 2
set.seed(41)
res <- c()
for (i in 1:nsim) {
    smp <- GRP( thePs, theN, BSDesign )
    w   <- anopa(frm, smp )
    res <- c(res, if(summarize(w)[2,4]<.05) 1 else 0)
}
typeI <- mean(res)
cat( "Design BxBxB, testing B: ", typeI, "\n")

# tolerance is large as the number of simulations is small
expect_equal( typeI, .05, tolerance = 0.035)
```



## Simulation with three factors, within design

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
frm      <- cbind(s.repue.early.1,  s.ajun.early.1,  s.repue.middle.1,
            s.ajun.middle.1, s.repue.late.1,   s.ajun.late.1,   s.repue.early.2, 
            s.ajun.early.2,  s.repue.middle.2, s.ajun.middle.2, s.repue.late.2,
            s.ajun.late.2,   s.repue.early.3,  s.ajun.early.3,  s.repue.middle.3,
            s.ajun.middle.3, s.repue.late.3,   s.ajun.late.3,   s.repue.early.4, 
            s.ajun.early.4,  s.repue.middle.4, s.ajun.middle.4, s.repue.late.4,
            s.ajun.late.4 ) ~ .
WSDesign <- list(eta = c("repue","ajun"), grp = c("early","middle","late"), a = c("1","2","3","4"))
thePs    <- rep(0.3, 24)

# test type-I error rate when no effect as is the case for factor 2
set.seed(43)
res <- c()
for (i in 1:nsim) {
    smp <- GRP( thePs, theN, NULL, WSDesign )
    w   <- anopa(frm, smp, WSFactors = c("e(2)","g(3)", "a(4)") )
    res <- c(res, if(summarize(w)[2,4]<.05) 1 else 0)
}
typeI <- mean(res)
cat( "Design WxWxW, testing W: ", typeI, "\n")

# tolerance is large as the number of simulations is small
expect_equal( typeI, .05, tolerance = 0.035)
```



## Simulation with three factors, mixed design, testing within

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
frm      <- cbind(s.repue.early,  s.ajun.early,  s.repue.middle,
s.ajun.middle, s.repue.late,  s.ajun.late ) ~ a
BSDesign <- list( a = c("1","2","3","4") )
WSDesign <- list(eta = c("repue","ajun"), grp = c("early","middle","late") )
thePs    <- rep(0.3, 24)

# test type-I error rate when no effect as is the case for factor 2
set.seed(43)
res <- c()
for (i in 1:nsim) {
    smp <- GRP( thePs, theN, BSDesign, WSDesign )
    w   <- anopa(frm, smp, WSFactors = c("e(2)","g(3)") )
    res <- c(res, if(summarize(w)[1,4]<.05) 1 else 0)
}
typeI <- mean(res)
cat( "Design BxWxW, testing W: ", typeI, "\n")

# tolerance is large as the number of simulations is small
expect_equal( typeI, .05, tolerance = 0.035)
```


## Simulation with three factors, mixed design, testing between

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
frm      <- cbind(s.repue.early,  s.ajun.early,  s.repue.middle,
s.ajun.middle, s.repue.late,  s.ajun.late ) ~ a
BSDesign <- list( a = c("1","2","3","4") )
WSDesign <- list(eta = c("repue","ajun"), grp = c("early","middle","late") )
thePs    <- rep(0.3, 24)

# test type-I error rate when no effect as is the case for factor 2
set.seed(42)
res <- c()
for (i in 1:nsim) {
    smp <- GRP( thePs, theN, BSDesign, WSDesign )
    w   <- anopa(frm, smp, WSFactors = c("e(2)","g(3)") )
    res <- c(res, if(summarize(w)[3,4]<.05) 1 else 0)
}
typeI <- mean(res)
cat( "Design BxWxW, testing B: ", typeI, "\n")

# tolerance is large as the number of simulations is small
expect_equal( typeI, .05, tolerance = 0.035)
```


# The end

Let's restore the warnings and messages before leaving:

```{r}
options("ANOPA.feedback" = 'all')
```




